Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-08-06 volunteer-clouds-ieee-ci-si.tex
Fixing typos, added new reference, changed figure,  and comments for @JJ and @mariosky

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream technology in the
same way the cloud technologies in general have. The lack of general purpose and open source tools
with analytics built in could be one of the reasons why this is so, as
well as the lack of proofs of concept. That is why in this paper we
present NodIO, a client-server architecture for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser. The architecture, along with some initial results on the expected performance and scaling
behavior is presented. We also use the algorithm data to  %[Pedro] I don't understand what "the algorithm analytics" stand for???
% data - JJ
find out what kind of patterns the user behaviour follow and try to fit them to some statistical distribution.
\end{abstract}

@@

\IEEEPARstart{N}{owadays} there is a big and increasing number of
distributed evolutionary computation software frameworks available on
the market, usually as open source systems. Most of these systems have
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, second,
they assume that the nodes that are going to run the system are under
your control or at least you have access to them.

@@

Some time ago, our group faced the problem of diminishing funds for
buying new     % [Pedro] This sentence is so so long. I would split it
               % in two sentences. [JJ] Done.
hardware. This was aggravated by the increasing maintance costs and extended downtime resulting from the continuous failures of existing clusters.
%[Mario] Otra opción: Some time ago, our group faced the problem of diminishing funds 
%for buying new hardware. This was aggravated by the increasing maintance costs 
%resulting from the continuous failures of existing clusters.
% [JJ] Done 
Considering this, we leveraged our experience in the design of web
applications, including JavaScript, since their inception and other
volunteer and 
unconventional distributed evolutionary computing systems, and we 
decided to designd and relase a new free framework that would allow anyone to
create a volunteer distributed EC experiment using cloud resources as
servers and browsers as clients.

These systems must solve a series of issues before even starting to
measure their performance; these issues are related to security as well as
safety, which introduces a {\em social} aspect to the design of a
computing system that makes it a {\em techno-social system}
\cite{vespignani2009predicting}. In this paper our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page that runs the system
determine its performance and how it can be modeled and, eventually,
predicted.

The rest of the paper is organized as follows: Next we present the
state of the art (Section \ref{sec:soa}) in web-based distributed 
computation systems along with attempts to predict and model its behavior. 
Then, a detailed description of the proposed system is presented 
in Section \ref{sec:description}, followed by the experiments and 
obtained results in Section \ref{sec:experiments}.
Finally, conclusions and future work are presented in Section \ref{sec:conclusion}.

@@

Recent works propose the use of cloud computing services for the components of 
client-server distributed EAs. Cloud resources 
can reduce operational costs by outsourcing both hardware and software maintenance 
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what 
is available in most laboratories, allowing researches to 
scale algorithms at reasonable costs. Researchers have proposed the use of 
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards} 
and DropBox~\cite{mericloud}. The use of JavaScript to implement server side components, %FERGU: he puesto mejor el de revista en lugar del del CEC
gives software developers certain advantages in the cloud paradigm. JavaScript is a first-classe citizen
in most Platforms as a Service (PaaS) as many support the Node.js framework \cite{wood13:nodejs:paas}. Also many web services %FERGU: he validado esta frase con un footnote, pero si quieres lo paso a referencias
have native JavaScript libraries and use JSON for data interchange. Another advantage is that 
developers can use the same software tools for both client and server side development. 
Besides the advantages to deal with the same language in the full stack as previously mentioned, 
this also facilitates the interoperability of the exchanged data types.
%Ayuda con la mini conclusion FERGU: mira la última frase de arriba

In the work presented in this paper, JavaScript has been used throughout the full 
stack, which has the advantage of using the same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library NodEO \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time
when writing code and leverages %FERGU: párrafo en pasado "was used" y
                                %presente "saves" 
% done - JJ
expertise acquired in the library itself. %FERGU: tienes info de LoC
                                %usadas para la sección experimental y
                                %justificar esta frase?
% Es el mismo código - JJ
%The proposed software has the advantage of full stack JavaScript
%development..
%FERGU: añadido el siguiente párrafo y referencia para justificar el uso de JS
Javascript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}. 
According to \cite{swanresearch2015}, the benefits of using this kind of paradigm in 
metaheuristics optimization allows an improvement in communicability, 
reproducibility, interoperability, automated assembly, knowledge engineering 
and efficiency.

@@

% [Pedro] desde el abstract no se había nombrado "NodIO". 
% Creo que en este párrafo debería introducirse, principalmente el nombre.
% Añado una frase presentándolo:
% [Mario] Nota: se menciona también en el segundo párrafo de la
% Introducción.
% OK [JJ]
In this sense, in this paper we propose {\sf NodIO}, a cloud or metal based 
volunteer computing system, whose architecture
has been developed using JavaScript on both the client and server sides.
All parts of the system are free and available with a free
license from \url{https://github.com/JJ/splash-volunteer}.
Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several remote %remote requests suena raro, propongo algo como: remote procedures o web interfaces   
  requests for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. The information
  is of two kinds: {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
    GET}ing a random chromosome from it, and {\em information} related
  to performance and state of the experiments. It also peforms logging
  duties, but is basically a very lightweight and hign performance
  data store \cite{jj:idc:lowcost}. 
  %The information is of two kinds: Problem Based y ¿Cual otra?
  %¿Logging? No se entiende bien. ¿Podría ser problem domain?   
% info. Añadido [JJ]
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found. 
\item A client that includes the evolutionary algorithm as 
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with random
  population and sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request). Apart from
  these constants, the algorithm can be run in many different ways:
  stopping when solution is found or, as we have done in this paper,
  using WebWorkers, which run in the background even if the webpage or
  browser tab with the experiment is not currently on focus.  
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture.
%
\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/system.pdf}
\caption{Description of the proposed system. Clients execute a JS EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one from the server. }  
\label{fig:system}
\end{figure}

The researcher only has to change a function to solve a different
problem, in this case the classical Trap function \cite{Ackley1987} has been 
used. JavaScript is a functional language and declaring a different
function and handing at the creation of the algorithm object, called
%FERGU: poner algo tras el handling: handling it?
%handing - JJ
{\tt Classic} is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in NodeJS and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a single thread
  allows the service of many requests. In fact, a limit with the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability.
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system or even browser requirement: anyone
  visiting the page, even from mobile devices, can load the algorithm. %Fergu: puesto lo de mobile devices para venderlo aún mejor
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through autonomous
  operation of every individual island and no synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server.
\item Since the algorithm is run on the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources that the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which means that operation is totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL and available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is bigger than what you would obtain in
  your own desktop system. If that is not done, it is a pointless
  academic exercise.
\end{itemize}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. They only compare on
runs in which solution has been found.}
\label{fig:baseline}
\end{figure}
First we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using {\sf NodEO},
the node.js evolutionary algorithm library. This experiment tries to
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$, a population size equal 512, which was found in previous
experiments to be the correct one for finding the solution 95\% of the
time. The algorithm was run until the solution (a string with all
ones) was found or five millions evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46
seconds$. Results for these two experiments 
are shown in Figure \ref{fig:baseline}.

These two experiments establish first a baseline result, which will
depend on the population. The volunteer computing experiments that we
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
destktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions.

>>>>

\begin{abstract}
Volunteer computing is used routinely in science and industry, but it
still has a long way to go if it is going to become a mainstream technology in the
same way the cloud technologies in general have. The lack of general purpose and open source tools
with analytics built in could be one of the reasons why this is so, as
well as the lack of proofs of concept. That is why in this paper we
present NodIO, a client-server architecture for distributed
evolutionary algorithms, written entirely in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser. The architecture, along with some initial results on the expected performance and scaling
behavior is presented. We also use the algorithm data to  %[Pedro] I don't understand what "the algorithm analytics" stand for???
% data - JJ
find out what kind of patterns the user behavior follow and try to fit them to some statistical distribution.
\end{abstract}

@@

\IEEEPARstart{N}{owadays} there is a big and increasing number of
distributed evolutionary computation software frameworks available on
the market, usually as open source systems \cite{Parejo12Survey}. Most of these systems have %FERGU: añadida referencia
two features: they are {\em desktop} applications, that is, they must
be compiled or run directly on the operating system layer and, second,
they assume that the nodes that are going to run the system are under
your control or at least you have access to them.

@@

Some time ago, our group faced the problem of diminishing funds for
buying new     % [Pedro] This sentence is so so long. I would split it
               % in two sentences. [JJ] Done.
hardware. This was aggravated by the increasing maintenance costs and extended downtime resulting from the continuous failures of existing clusters.
%[Mario] Otra opción: Some time ago, our group faced the problem of diminishing funds 
%for buying new hardware. This was aggravated by the increasing maintance costs 
%resulting from the continuous failures of existing clusters.
% [JJ] Done 
Considering this, we leveraged our experience in the design of web
applications, including JavaScript, since their inception and other
volunteer and 
unconventional distributed evolutionary computing systems, and we 
decided to design and release a new free framework that would allow anyone to
create a volunteer distributed EC experiment using cloud resources as
servers and browsers as clients.

This kind of systems must solve a series of issues before even starting to %FERGU: cambiado el "these systems" porque parecía que hacía referencia a "los anteriores" 
measure their performance; these issues are related to security as well as
safety, which introduces a {\em social} aspect to the design of a
computing system that makes it a {\em techno-social system}
\cite{vespignani2009predicting}. In this paper our intention is to
present the {\sf NodIO} framework and analyze it from two different
angles: the purely technical angle, including the decisions that went
into its features, and the social angle, to account on how the persons
that voluntarily choose to enter the web page that runs the system
determine its performance and how it can be modeled and, eventually,
predicted.

The rest of the paper is organized as follows: Next we present the
state of the art (Section \ref{sec:soa}) in web-based distributed 
computation systems along with attempts to predict and model its behavior. %FERGU: El IEEE es americano
Then, a detailed description of the proposed system is presented 
in Section \ref{sec:description}, followed by the experiments and 
obtained results in Section \ref{sec:experiments}.
Finally, conclusions and future work are presented in Section \ref{sec:conclusion}.

@@

Recent works propose the use of cloud computing services for the components of 
client-server distributed EAs. Cloud resources 
can reduce operational costs by outsourcing both hardware and software maintenance 
to the cloud provider. Another advantage is that they enable the provisioning of computing resources beyond what 
is available in most laboratories, allowing researches to 
scale algorithms at reasonable costs. Researchers have proposed the use of 
public clouds such as Amazon EC2 \cite{CloudScale}, Google App Engine\cite{di2013towards} 
and DropBox~\cite{mericloud}. The use of JavaScript to implement server side components, %FERGU: he puesto mejor el de revista en lugar del del CEC
gives software developers certain advantages in the cloud paradigm. JavaScript is a first-classe citizen
in most Platforms as a Service (PaaS) as many support the Node.js framework \cite{wood13:nodejs:paas}. Also many web services 
have native JavaScript libraries and use JSON for data interchange. Another advantage is that 
developers can use the same software tools for both client and server side development. 
Besides the advantages to deal with the same language in the full stack as previously mentioned, 
this also facilitates the interoperability of the exchanged data types.
%Ayuda con la mini conclusion FERGU: mira la última frase de arriba

In the work presented in this paper, JavaScript has been used throughout the full 
stack, which has the advantage of using the same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library NodEO \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time
when writing code and leverages %FERGU: qué diferencias hay entre NodIO y NodEO??
expertise acquired in the library itself. %FERGU: tienes info de LoC
                                %usadas para la sección experimental y
                                %justificar esta frase?
% Es el mismo código - JJ
%The proposed software has the advantage of full stack JavaScript
%development..

Javascript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}. 
According to \cite{swanresearch2015}, the benefits of using this kind of paradigm in 
metaheuristics optimization allows an improvement in communicability, 
reproducibility, interoperability, automated assembly, knowledge engineering 
and efficiency.

@@

% [Pedro] desde el abstract no se había nombrado "NodIO". 
% Creo que en este párrafo debería introducirse, principalmente el nombre.
% Añado una frase presentándolo:
% [Mario] Nota: se menciona también en el segundo párrafo de la
% Introducción.
% OK [JJ]
In this sense, in this paper we propose {\sf NodIO}, a cloud or metal based 
volunteer computing system, whose architecture
has been developed using JavaScript on both the client and server sides.
All parts of the system are free and available with a free
license from \url{https://github.com/JJ/splash-volunteer}.
Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several remote %remote requests suena raro, propongo algo como: remote procedures o web interfaces   
  requests for storing and retrieving information (the CRUD cycle:
  create, request, update and delete) from the server. The information
  is of two kinds: {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
    GET}ing a random chromosome from it, and {\em information} related
  to performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}. 
  %The information is of two kinds: Problem Based y ¿Cual otra?
  %¿Logging? No se entiende bien. ¿Podría ser problem domain?   
% info. Añadido [JJ]
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found. 
\item A client that includes the evolutionary algorithm as 
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with random
  population and sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request). Apart from
  these constants, the algorithm can be run in many different ways:
  stopping when solution is found or, as we have done in this paper,
  using WebWorkers, which run in the background even if the webpage or
  browser tab with the experiment is not currently on focus.  
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different web technologies, such as JQuery or chart.js have
been used to build the elements of the system. 
%FERGU: Mario, explica en el texto de abajo qué es winston en la figura también. 
%Vamos, que todo lo que salga en la figura tiene que ir en el texto en alguna parte :)
%
\begin{figure}[!t]
\centering
\includegraphics[width=4in]{img/Algorithm.pdf}
\caption{Description of the proposed system. Clients execute a JS EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one from the server. }  
\label{fig:system}
\end{figure}

The researcher only has to change a function to solve a different
problem, in this case the classical Trap function \cite{Ackley1987} has been 
used. JavaScript is a functional language and declaring a different
function and handing at the creation of the algorithm object, called
{\tt Classic} is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in NodeJS and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a single thread
  allows the service of many requests. In fact, a limit with the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability.
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system or even browser requirement: anyone
  visiting the page, even from mobile devices, can load the algorithm. 
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through autonomous
  operation of every individual island and no synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server.
\item Since the algorithm is run on the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources that the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which means that operation is totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL and available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is bigger than what you would obtain in
  your own desktop system. If that is not done, it is a pointless
  academic exercise.
\end{itemize}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{img/baseline-times.png}
\caption{Comparison of time to solution for the baseline system, using a {\tt
    node.js} client, for different population sizes. They only compare on
runs in which solution has been found.}
\label{fig:baseline}
\end{figure}
First we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO, 
%FERGU: es NodEO o NodIO? Quito el \sf, por cierto, que queda feo
%Por qué usamos NodEO aquí y no NodIO, vamos a compararlos? 
%Qué diferencias hay entre los dos? (ponerlo en el SOA)
the node.js evolutionary algorithm library. This experiment tries to 
%FERGU: quitar el "the node.js evolutionary algorithm library", si ya lo hemos dicho
%o explicarlo mejor: "the basic js library we have used as base to built NodIO", por ejemplo
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$, a population size equal 512, which was found in previous
experiments to be the correct one for finding the solution 95\% of the
%Fergu: cita a esos experimentos?
time. The algorithm was run until the solution (a string with all
ones) was found or five millions evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with success rate upgraded to 100\% and average duration $3.46
seconds$. Results for these two experiments 
are shown in Figure \ref{fig:baseline}.

These two experiments establish first a baseline result, which will
depend on the population. The volunteer computing experiments that we
will describe next do not and can not have the same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions.
